---
title: "Getting started with fastsemR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fastsemR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("fastsemR", quietly = TRUE) &&
         isTRUE(fastsemR:::.fastsem_env$loaded)
)
library(fastsemR)
```

fastsemR wraps the [fastsem](https://github.com/lf-araujo/fastsem) compiled
SEM engine.  Unlike lavaan it uses a pre-compiled Nim binary, which keeps R
memory management out of the hot loop and allows optional GPU acceleration.
This vignette walks through the main use cases using only base-R datasets.

## Installation and setup

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("lf-araujo/fastsemR")

# The binary is downloaded automatically on first load.
# To upgrade or to point at a local build:
library(fastsemR)
fastsem_update()                              # download latest release
fastsem_load("~/fastsem/libfastsem_r.so")    # or use a local build
```

Once the library is loaded, `fastsem_fit()` is available.

---

## 1. Path model (observed variables)

We regress fuel economy (`mpg`) on horsepower (`hp`) and weight (`wt`),
allowing the two predictors to covary.  Data are scaled so default starting
values work well.

```{r path-model}
df <- as.data.frame(scale(mtcars[, c("mpg", "hp", "wt")]))

syntax_path <- "
  # Structural part
  mpg ~ hp + wt
  # Predictor covariance
  hp ~~ wt
  # Residual / total variances
  hp ~~ hp
  wt ~~ wt
  mpg ~~ mpg
  # Intercepts (needed for FIML even on complete data)
  mpg ~ 1
  hp  ~ 1
  wt  ~ 1
"

res_path <- fastsem_fit(syntax_path, df)
print_fastsem(res_path)
```

The saturated path model has `df = 0`, so `chi2 = 0` and fit indices are
undefined.  Adding a constraint (e.g. fixing the `hp → mpg` path to zero)
would give `df = 1` and a testable chi-square.

---

## 2. Confirmatory factor analysis

A single factor `g` is measured by three positively correlated iris
variables.  We fix the latent variance to 1 (`g ~~ 1*g`) to set the scale
instead of fixing a loading.

```{r cfa}
df_iris <- as.data.frame(scale(iris[, c(1, 3, 4)]))
names(df_iris) <- c("sl", "pl", "pw")

syntax_cfa <- "
  g =~ sl + pl + pw
  g ~~ 1*g
  sl ~~ sl
  pl ~~ pl
  pw ~~ pw
"

res_cfa <- fastsem_fit(syntax_cfa, df_iris)
print_fastsem(res_cfa)
```

The standardised estimates (StdAll) are printed automatically when available.
All three loadings should be close to 0.85–1.00 for the strongly correlated
iris variables.

---

## 3. FIML with missing data

fastsem automatically activates FIML when `NA` values are present.  The
estimator label in the output switches from `"ML"` to `"FIML"`.

```{r fiml}
set.seed(99)
df_miss <- as.data.frame(scale(mtcars[, c("mpg", "hp", "wt")]))
df_miss$mpg[sample(32, 5)] <- NA   # 15% missing on mpg
df_miss$wt[sample(32, 3)]  <- NA   # 9% missing on wt

res_fiml <- fastsem_fit(syntax_path, df_miss)
cat("Estimator:", res_fiml$estimator, "\n")
cat("hp->mpg:", round(res_fiml$estimates[res_fiml$paramNames == "hp->mpg"], 3), "\n")
cat("wt->mpg:", round(res_fiml$estimates[res_fiml$paramNames == "wt->mpg"], 3), "\n")
```

Estimates with partial missingness should be close to the complete-data
values above because the data were simulated without a missingness mechanism.

---

## 4. Parameter labels, equality constraints, and bounds

### 4a. Labels and the `start()` token

Label a path `b1` so that its estimate appears in the output under that
name and can be referenced in derived parameters.

```{r labels}
syntax_lab <- "
  mpg ~ b1*hp + b2*wt
  hp ~~ wt
  hp ~~ hp;  wt ~~ wt;  mpg ~~ mpg
  mpg ~ 1;   hp ~ 1;    wt ~ 1
"
res_lab <- fastsem_fit(syntax_lab, df)
cat("b1 (hp->mpg):", round(res_lab$estimates[res_lab$paramNames == "b1"], 3), "\n")
cat("b2 (wt->mpg):", round(res_lab$estimates[res_lab$paramNames == "b2"], 3), "\n")
```

### 4b. Equality constraints

Two paths with the same label are constrained to be equal.  Here we force
`hp` and `wt` to have the same slope onto `mpg`:

```{r equality}
syntax_eq <- "
  mpg ~ eq*hp + eq*wt
  hp ~~ wt
  hp ~~ hp;  wt ~~ wt;  mpg ~~ mpg
  mpg ~ 1;   hp ~ 1;    wt ~ 1
"
res_eq <- fastsem_fit(syntax_eq, df)
cat("Constrained slope (eq):", round(res_eq$estimates[res_eq$paramNames == "eq"], 3), "\n")
cat("df (1 constraint):", res_eq$df, "\n")
cat("chi2:", round(res_eq$chi2, 3), "  p =", round(res_eq$pvalue, 3), "\n")
```

### 4c. Bounds

Bound the `hp → mpg` coefficient to be negative (upper bound = 0):

```{r bounds}
syntax_bnd <- "
  mpg ~ b1*hp + b2*wt
  hp ~~ wt
  hp ~~ hp;  wt ~~ wt;  mpg ~~ mpg
  mpg ~ 1;   hp ~ 1;    wt ~ 1
  b1 < 0
"
res_bnd <- fastsem_fit(syntax_bnd, df)
cat("b1 with upper bound 0:", round(res_bnd$estimates[res_bnd$paramNames == "b1"], 4), "\n")
```

---

## 5. Derived parameters (indirect effects)

The `:=` operator computes a derived quantity at the optimum and applies the
delta method to obtain its SE.  Here we test a simple mediation:
`hp → wt → mpg`.

```{r indirect}
syntax_med <- "
  wt  ~ a*hp
  mpg ~ b*wt + c_prime*hp
  hp ~~ hp;  wt ~~ wt;  mpg ~~ mpg
  hp ~ 1;    wt ~ 1;    mpg ~ 1
  ab := a * b
"
res_med <- fastsem_fit(syntax_med, df)

idx_ab <- which(res_med$paramNames == "ab")
cat(sprintf("Indirect (hp->wt->mpg):  est = %6.4f,  SE = %6.4f,  z = %5.2f\n",
            res_med$estimates[idx_ab],
            res_med$se[idx_ab],
            res_med$z[idx_ab]))
```

The z-statistic tests whether the indirect effect differs from zero under
the delta-method normal approximation.

---

## 6. Comparison with OpenMx

When working with `umxRAM()` models, `run_fastsem()` handles translation and
inject-back automatically.  The estimates should match OpenMx's ML
estimates closely.

```{r compare-omx, eval=requireNamespace("umx", quietly=TRUE) && isTRUE(fastsemR:::.fastsem_env$loaded)}
library(umx)
umx_set_auto_run(FALSE)

m_fs <- umxRAM("path_fs",
  umxPath(from = c("hp","wt"), to = "mpg"),
  umxPath(cov   = c("hp","wt")),
  umxPath(var   = c("hp","wt","mpg")),
  umxPath(means = c("hp","wt","mpg")),
  data = df
)
m_fs <- run_fastsem(m_fs)

umx_set_auto_run(TRUE)
m_omx <- umxRAM("path_omx",
  umxPath(from = c("hp","wt"), to = "mpg"),
  umxPath(cov   = c("hp","wt")),
  umxPath(var   = c("hp","wt","mpg")),
  umxPath(means = c("hp","wt","mpg")),
  data = df
)
umx_set_auto_run(FALSE)

cat("\nfastsem vs OpenMx (hp->mpg, wt->mpg):\n")
print(round(rbind(
  fastsem = omxGetParameters(m_fs)[c("hp_to_mpg","wt_to_mpg")],
  OpenMx  = omxGetParameters(m_omx)[c("hp_to_mpg","wt_to_mpg")]
), 4))
```

---

## 7. Multi-group CFA

Multi-group models are handled by `run_fastsem()` automatically when the
`MxModel` contains submodels with separate data.  Here we fit a configural
CFA on the Holzinger-Swineford visual subscale across two schools.

```{r multigroup, eval=requireNamespace("umx", quietly=TRUE) && requireNamespace("lavaan", quietly=TRUE) && isTRUE(fastsemR:::.fastsem_env$loaded)}
library(lavaan, warn.conflicts = FALSE)
hs <- HolzingerSwineford1939

mg_p <- umxRAM("mg_p",
  umxPath(from = "g", to = c("x1","x2","x3")),
  umxPath(v1m0 = "g"),
  umxPath(var   = c("x1","x2","x3")),
  umxPath(means = c("x1","x2","x3")),
  data = hs[hs$school == "Pasteur",     c("x1","x2","x3")]
)
mg_gw <- umxRAM("mg_gw",
  umxPath(from = "g", to = c("x1","x2","x3")),
  umxPath(v1m0 = "g"),
  umxPath(var   = c("x1","x2","x3")),
  umxPath(means = c("x1","x2","x3")),
  data = hs[hs$school == "Grant-White", c("x1","x2","x3")]
)

r <- run_fastsem(mxModel("HS_mg", mg_p, mg_gw))

cat("Pasteur loadings:\n")
print(round(omxGetParameters(r@submodels$mg_p)[1:3], 3))
cat("\nGrant-White loadings:\n")
print(round(omxGetParameters(r@submodels$mg_gw)[1:3], 3))
```

The two schools produce similar but not identical loadings, consistent with
approximate (not strict) metric invariance for this subscale.

---

## Next steps

- See `vignette("behavior-genetics")` for ACE twin models and bivariate
  Cholesky decompositions.
- See `?fastsem_fit` for the full lavaan syntax reference including ordinal
  variables, cluster-robust SEs, and definition variables.
