---
title: "State Space Models with fastsemR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State Space Models with fastsemR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("fastsemR", quietly = TRUE) &&
         isTRUE(fastsemR:::.fastsem_env$loaded)
)
library(fastsemR)
set.seed(42)
```

State space models (SSMs) are longitudinal models in which a latent process
evolves over time according to a **transition equation**, and is observed with
error via an **observation equation**:

$$
\mathbf{x}_t = \mathbf{A}\,\mathbf{x}_{t-1} + \mathbf{B}\,\mathbf{u}_t + \mathbf{w}_t,
\qquad \mathbf{w}_t \sim \mathcal{N}(\mathbf{0}, \mathbf{Q})
$$
$$
\mathbf{y}_t = \mathbf{C}\,\mathbf{x}_t + \mathbf{D}\,\mathbf{u}_t + \mathbf{v}_t,
\qquad \mathbf{v}_t \sim \mathcal{N}(\mathbf{0}, \mathbf{R})
$$

Parameters are estimated via the **Kalman filter** log-likelihood, maximised
with the Adam optimiser using an analytical gradient (sensitivity
propagation).  Standard errors come from the numerical Hessian of −2 ln *L*.

**Computational backend.**  The Kalman objective is evaluated on the GPU
(OpenCL float32) when the model fits within nS ≤ 8, nO ≤ 8, nI ≤ 8 — which
covers all standard SSM models and the longitudinal GWAS path (nI = 1).  The
gradient always runs on CPU (Weave-parallel analytical sensitivity).

fastsemR exposes the full SSM path through `fastsem_fit_ssm()`, which accepts
a formula string in fastsem's SSM syntax alongside a long-format data frame.

---

## Data format

SSM data must be in **long format**: one row per person–wave combination with
at minimum a person ID column, a time/wave column, and the observed outcome
columns.

```{r sim-data}
# Simulate a simple AR(1) latent process observed through two indicators
n_persons <- 150
n_waves   <- 5

sim_ssm <- function(n, T, a = 0.7, sigma_q = 0.5, sigma_r = 0.3) {
  rows <- vector("list", n * T)
  k    <- 1L
  for (i in seq_len(n)) {
    x <- rnorm(1, 0, sqrt(sigma_q^2 / (1 - a^2)))  # stationary init
    for (t in seq_len(T)) {
      x    <- a * x + rnorm(1, 0, sigma_q)
      y1   <- x + rnorm(1, 0, sigma_r)
      y2   <- x + rnorm(1, 0, sigma_r * 1.2)
      rows[[k]] <- data.frame(pid = i, wave = t, y1 = y1, y2 = y2)
      k <- k + 1L
    }
  }
  do.call(rbind, rows)
}

df_ssm <- sim_ssm(n_persons, n_waves)
head(df_ssm, 8)
```

---

## 1. Univariate AR(1) latent process

A single latent state `x` follows an AR(1) process.  Two observed variables
`y1` and `y2` each load on `x` with fixed unit loadings — so `x` is defined
on the same scale as the observations.

```{r ar1}
syntax_ar1 <- "
  id:   pid
  time: wave

  # State transition (A matrix)
  x ~> x

  # Process noise (diagonal Q)
  x ~~ x

  # Observation model: state x is manifest in y1 and y2 (fixed unit loadings)
  x =~ 1*y1 + 1*y2

  # Measurement error (diagonal R)
  y1 ~~ y1
  y2 ~~ y2

  # Initial state covariance
  x ~~ x_init
"

res_ar1 <- fastsem_fit_ssm(syntax_ar1, df_ssm)
print_fastsem(res_ar1)
```

The output reports:

- **transition** — the AR(1) coefficient A[x,x] (should be near 0.70)
- **proc.noise** — process noise variance Q[x,x] (near 0.25 = 0.5²)
- **meas.error** — measurement error variances R[y1,y1] and R[y2,y2]
- **init.cov** — initial state variance P0[x,x]
- **Model Fit** — −2 ln L, AIC, BIC

---

## 2. Bivariate cross-lagged process

Two latent states `x` and `y` can influence each other over time.  This is
the SSM analogue of a cross-lagged panel model.

```{r clpm-sim}
# Simulate bivariate cross-lagged data
sim_bivariate <- function(n, T,
                          axx = 0.6, axy = 0.2,
                          ayx = 0.1, ayy = 0.5,
                          sigma_q = 0.4, sigma_r = 0.3) {
  rows <- vector("list", n * T)
  k <- 1L
  for (i in seq_len(n)) {
    xs <- rnorm(1, 0, 0.5); ys <- rnorm(1, 0, 0.5)
    for (t in seq_len(T)) {
      xs_new <- axx * xs + axy * ys + rnorm(1, 0, sigma_q)
      ys_new <- ayx * xs + ayy * ys + rnorm(1, 0, sigma_q)
      xs <- xs_new; ys <- ys_new
      rows[[k]] <- data.frame(
        pid = i, wave = t,
        ox = xs + rnorm(1, 0, sigma_r),
        oy = ys + rnorm(1, 0, sigma_r)
      )
      k <- k + 1L
    }
  }
  do.call(rbind, rows)
}

df_clpm <- sim_bivariate(n_persons, n_waves)
```

```{r clpm}
syntax_clpm <- "
  id:   pid
  time: wave

  # All four A-matrix elements (autoregressive + cross-lagged)
  x ~> x
  x ~> y
  y ~> x
  y ~> y

  # Process noise (diagonal Q — uncorrelated innovations)
  x ~~ x
  y ~~ y

  # Observations: one indicator per state
  x =~ 1*ox
  y =~ 1*oy

  # Measurement error
  ox ~~ ox
  oy ~~ oy

  # Initial covariances
  x ~~ x_init
  y ~~ y_init
"

res_clpm <- fastsem_fit_ssm(syntax_clpm, df_clpm)
print_fastsem(res_clpm)
```

The four transition parameters correspond to the A matrix:

| Parameter | Meaning         | True value |
|-----------|-----------------|------------|
| `x~>x`    | AR for x        | 0.60       |
| `y~>x`    | x predicts y    | 0.10       |
| `x~>y`    | y predicts x    | 0.20       |
| `y~>y`    | AR for y        | 0.50       |

---

## 3. Exogenous inputs (B and D matrices)

Exogenous variables — time-varying covariates or genetic dosages — enter
the model via the `~` operator:

- `state ~ coef * input` → **B matrix** (input affects latent state)
- `obs ~ coef * input` → **D matrix** (input affects observation directly)

```{r exog-sim}
# Simulate with a binary group covariate that shifts the latent level
sim_with_input <- function(n, T, b_input = 0.8, a = 0.65,
                           sigma_q = 0.4, sigma_r = 0.3) {
  rows <- vector("list", n * T)
  k <- 1L
  for (i in seq_len(n)) {
    grp <- rbinom(1, 1, 0.5)   # time-invariant covariate
    x   <- rnorm(1, 0, 0.5)
    for (t in seq_len(T)) {
      x <- a * x + b_input * grp + rnorm(1, 0, sigma_q)
      y <- x + rnorm(1, 0, sigma_r)
      rows[[k]] <- data.frame(pid = i, wave = t, y = y, grp = grp)
      k <- k + 1L
    }
  }
  do.call(rbind, rows)
}

df_input <- sim_with_input(n_persons, n_waves)
```

```{r exog}
syntax_input <- "
  id:   pid
  time: wave

  x ~> x
  x ~~ x
  x =~ 1*y
  y ~~ y
  x ~~ x_init

  # B matrix: group covariate shifts the latent state
  x ~ b_grp * grp
"

res_input <- fastsem_fit_ssm(syntax_input, df_input)
print_fastsem(res_input)
```

The parameter `b_grp` (type `input→state`) estimates how much the group
variable shifts the latent state at each wave.  The true value is 0.80.

---

## 4. Missing data

The Kalman filter handles missing observations naturally: at any wave where
`y` is `NA` the update step is skipped and the filter propagates the predicted
state.  No special syntax is needed — just leave `NA` in the data frame.

```{r missing}
df_miss <- df_ssm
# Introduce 20% missingness on y1
df_miss$y1[sample(nrow(df_miss), size = floor(0.2 * nrow(df_miss)))] <- NA

res_miss <- fastsem_fit_ssm(syntax_ar1, df_miss)
cat("AR coefficient with 20% missingness:",
    round(res_miss$estimates[res_miss$paramNames == "x~>x"], 3), "\n")
```

---

## 5. Fixing parameters

Use `1*` to fix a transition coefficient (e.g. a random walk where A = 1),
or any numeric prefix to fix at that value:

```{r fixed}
syntax_rw <- "
  id:   pid
  time: wave

  # Random walk: A[x,x] = 1 (fixed)
  1*x ~> x

  x ~~ x
  x =~ 1*y1
  y1 ~~ y1
  x ~~ x_init
"

res_rw <- fastsem_fit_ssm(syntax_rw, df_ssm[df_ssm$wave <= 3, ])
cat("Process noise (random walk):",
    round(res_rw$estimates[res_rw$paramNames == "x~~x"], 4), "\n")
```

---

## 6. Longitudinal GWAS sweep

When a `geno:` directive is present alongside the SSM syntax, fastsem runs a
**per-SNP Kalman sweep**: for each variant the dosage is injected as a
time-invariant exogenous input and the model is re-fitted with a warm start.
This tests whether a SNP shifts the level of the latent state.

Two genotype formats are supported:

- **PLINK binary** (`.bed` / `.bim` / `.fam`): give the prefix without extension.
- **Plain-text VCF** (`.vcf`): `GT` hard-calls or `DS` dosage field; bgzipped
  files (`.vcf.gz`) are not yet supported.

The model must declare the SNP path explicitly:

```r
# /data/longitudinal_cohort.csv
id:   pid
time: wave
geno: /data/plink_prefix     # PLINK binary — OR — /data/genotypes.vcf for VCF
sample_col: pid               # CSV column to match genomic sample IDs

x ~> x
x ~~ x
x =~ 1*y1;  y1 ~~ y1
x ~~ x_init

# One free parameter per SNP: B[x, snp]
x ~ b_snp * snp
```

Output (tab-separated to stdout) — **all free parameters** are reported per SNP:

```
CHR  POS  SNP  A1  A2  b_snp  SE_b_snp  x~>x  SE_x~>x  Q[x]  SE_Q[x]  R[y1]  SE_R[y1]  ...  Z  P  N_PERSONS
1    4137824  rs4137824  A  G  0.392  0.041  0.651  0.028  0.091  0.018  0.040  0.008  ...  9.56  1.2e-21  300
```

The columns are:

- **Locus**: `CHR`, `POS`, `SNP`, `A1`, `A2`
- **Per-parameter pairs**: `<name>` and `SE_<name>` for every free parameter in the model (SNP B-coefficient, AR coefficients, process noise, measurement error, initial covariance)
- **`Z`** and **`P`**: Wald statistic and two-sided p-value for the SNP parameter only
- **`N_PERSONS`**: number of genotyped persons for this SNP

After the sweep, fastsem re-fits the **null model** (no SNP input) and prints a full
parameter report to `stderr` — this gives you clean baseline estimates for the latent
process unconfounded by any single SNP.

Both PLINK binary (`.bed`/`.bim`/`.fam`) and plain-text VCF files are
supported.  The first SNP uses a cold start (800 Adam iterations); subsequent
SNPs warm-start from the previous result (200 iterations), making genome-chip
sweeps (~500k variants) tractable.

```r
# Run a longitudinal GWAS sweep from R
res_gwas <- fastsem_fit_ssm(syntax_gwas, df_long,
                             geno = "/data/plink_prefix",
                             sample_col = "pid")
# res_gwas is a data.frame: CHR POS SNP A1 A2 <param cols> SE_<param cols> Z P N_PERSONS
head(res_gwas)
```

---

## Summary of SSM syntax

| Syntax              | Matrix | Meaning                                      |
|---------------------|--------|----------------------------------------------|
| `x ~> x`            | A      | Autoregressive coefficient                   |
| `x ~> y`            | A      | Cross-lagged: x at t−1 predicts y at t       |
| `1*x ~> x`          | A      | Fixed transition coefficient (= 1)           |
| `x ~~ x`            | Q      | Process noise variance (log-parameterised)   |
| `x ~~ z`            | Q      | Process noise covariance (tanh form)         |
| `x =~ y1`           | C      | Observation loading (state x → obs y1)       |
| `x =~ 1*y1`         | C      | Fixed unit loading                           |
| `x =~ 1*y1 + y2`    | C      | Multiple observations per state              |
| `y1 ~~ y1`          | R      | Measurement error variance                   |
| `x ~~ x_init`       | P0     | Initial state variance                       |
| `x ~ b*u`           | B      | Exogenous input → latent state               |
| `y1 ~ d*u`          | D      | Exogenous input → observation                |
| `id:   col`         | —      | Person ID column (required)                  |
| `time: col`         | —      | Wave / time column (required)                |
| `geno: path`        | —      | PLINK prefix or VCF for longitudinal GWAS    |
| `sample_col: col`   | —      | CSV column to align genomic sample IDs       |

---

## See also

- `vignette("getting-started")` — cross-sectional SEM, CFA, FIML, multi-group
- `vignette("behavior-genetics")` — ACE twin models and Cholesky decompositions
- `fastsem help ssm` — full SSM reference from the command line
- `fastsem help syntax` — complete syntax reference
